### 2015-05-14 notes.tex
initial commit

-smoothed analysis

--what is it? option when theory is far from practice, so good in the
learning setting. randomly perturb (``smooth'') part of a problem and
examine what happens on that smoothed instance. give a few classic
examples here; simplex, tsp.

--what assumptions does it make about the data? when is it useful to
apply?  when something about problem is not adversarial; i.e. may
arise randomly in practice.

--smoothed analysis in learning. still new; people unsure where to
apply it: lots of different things to smooth. smooth distribution or
smooth points? what does it mean to smooth distribution? don't
know. classically in SA there's one clear thing to smooth but in
learning we don't really know what's best yet. these 2 papers contain
different approaches. k-means is runtime-focused. smooths points and
shows that the runtime on the perturbed set is expected
polynomial. tensor decomp is accuracy focused. perturbs vectors and
shows that factorization works robustly with perturbations. these 2
approaches are just examples in the current lines of research. we
could also imagine doing [OTHER CREATIVE THING] In classic model,
algorithm performance is measured by runtime. In learning we have
another metric: accuracy. can use smoothing to get results on both
fronts; see the two examples below.

% maybe do the main result from the tensor paper instead and mention
% this result at a higher level. in the conclusion mention how the two
% papers take different avenues.

\section{$k$-means}
\subsection{Algorithm}
The $k$-means algorithm solves the \emph{clustering
  problem}, set as follows. Given a set $S$ of points in a (generally
high-dimensional) space $\mathbb{R}^d$, output a $k$-partition of the
set (a collection of $k$ ``clusters'') such that the elements in each
cluster are similar under a chosen metric. $k$-means is iterative:
first it arbitrarily chooses the $k$ cluster centers, then assigns each point in $S$ to the closest center. Next, it
loops over these clusters and computes the mean of all
points in the cluster. These means become the new cluster centers and
the assignment process repeats. The algorithm terminates when the
local improvements to cluster centers no longer affect the assignment
of points.

\subsection{$k$-means is Exponential}
In \cite{KMworstcase}, Vattani shows that an adversary can choose
initial cluster centers that will force $k$-means to iterate
$2^{\Omega(n)}$ times for $d \ge 2$. This construction is based on
[SOMETHING].

\subsection{Smoothed Analysis}
We give a summary of the argument in \cite{SAkmeans}, by Arthur,
Manthey, and Roeglin, that $k$-means has polynomial smoothed
complexity. The main result of this section is the following theorem.
\begin{theorem}
Fix a set $S^{\prime} \subset [0,1]^d$ of $n$ data points, and
independently perturb each point in $S^{\prime}$ by a Gaussian
distribution with mean 0 and variance $\sigma^2$. Label the new set
$S$. Then the expected running time of $k$-means on $S$ is bounded by
a polynomial in $n$ and $1/\sigma$.
\end{theorem}
We reason about the potential function \begin{equation}
\Psi = \sum_{x \in S}||x - c(x)||^2,
\end{equation} 
where $c(x)$ is the cluster center to which the point $x$ is assigned
in the current iteration. The proof relies on two lemmas: first, that
$\Psi$ is bounded by a polynomial after one iteration (define
iteration) of the algorithm, and second that $\Psi$ decreases by a
polynomial amount in each subsequent iteration (actually, sequences of
a few iterations).

\begin{lemma}
The potential after one iteration is bounded by a polynomial.
\end{lemma}
\begin{proof}
\end{proof}

### 2015-05-14 old.tex
initial commit

--- 2015-05-14 notes.tex
@@
+\documentclass[11pt]{article}
+
+\usepackage{amsthm, amssymb, amsmath}
+\usepackage{rotating}
+\usepackage{verbatim}
+\usepackage{enumerate}
+\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry}
+\usepackage{multicol} % Used for the two-column layout of the document
+\usepackage[hang, small,labelfont=bf,up,textfont=it,up]{caption}
+\usepackage{booktabs} % Horizontal rules in tables
+\usepackage{hyperref} % For hyperlinks in the PDF
+\usepackage[shortlabels]{enumitem}
+\usepackage{abstract} % Allows abstract customization
+\renewcommand{\abstractnamefont}{\normalfont\bfseries} % Set the ``Abstract'' text to bold
+\renewcommand{\abstracttextfont}{\normalfont\small\itshape} % Set the abstract itself to small italic text
+
+\usepackage{titlesec} % Allows customization of titles
+\titlelabel{\thetitle.\quad}
+%\renewcommand\thesection{\Roman{section}} % Roman numerals for the sections
+%\renewcommand\thesubsection{\Roman{subsection}} % Roman numerals for subsections
+%\titleformat{\section}[block]{\large\scshape\centering}{\thesection.}{1em}{} % Change the look of the section titles
+%\titleformat{\subsection}[block]{\large}{\thesubsection.}{1em}{} % Change the look of the section titles
+
+\usepackage{fancyhdr} % Headers and footers
+\pagestyle{fancy} % All pages have headers and footers
+\fancyhead{} % Blank out the default header
+\fancyfoot{} % Blank out the default footer
+\fancyhead[C]{May 2015} % Custom header text
+\fancyfoot[RO,LE]{\thepage} % Custom footer text
+
+\linespread{1.05}
+\bibliographystyle{plain}
+\newtheorem{lemma}{Lemma}
+\newtheorem{corollary}{Corollary}
+\theoremstyle{definition}
+\newtheorem{theorem}{Theorem}
+
+% --------------------
+% TITLE SECTION
+% --------------------
+
+\title{\vspace{-5mm}\fontsize{22pt}{10pt}\selectfont Smoothed Analysis in Learning: Tensors and $k$-Means}
+
+\author{
+\large
+\textsc{Hunter Lang} \\
+\normalsize MIT \\
+\normalsize \href{mailto:hjl@mit.edu}{\texttt{hjl@mit.edu}}\\
+\and
+\textsc{Carlos Cortez} \\
+\normalsize MIT \\
+\normalsize \href{mailto:cortezc@mit.edu}{\texttt{cortezc@mit.edu}}\\
+\vspace{-5mm}
+}
+\date{}
+
+\begin{document}
+\maketitle
+
+%--------------------
+% ABSTRACT
+%--------------------
+
+\begin{abstract}
+\noindent Most learning problems are hard in the worst-case, so much
+of the current research focuses on finding special cases with provably
+good time complexities, or polynomial-time approximation algorithms. But many algorithms (Simplex being the
+classical example) seem to consistently run fast
+in practice. Smoothed
+analysis combines worst-case and average-case analysis to better
+approximate real-world algorithm performance and close the gap between
+theoretical and practical results. Here we survey the application of
+smoothed analysis to $k$-means and tensor decomposition, and show that
+both problems admit algorithms with polynomial smoothed complexity, in line with empirical oberservations.
+\end{abstract}
+
+\section{Introduction}
 -smoothed analysis
@@
 learning setting. randomly perturb (``smooth'') part of a problem and
-examine what happens on that smoothed instance. give a few classic
-examples here; simplex, tsp.
+examine what happens on that smoothed instance.give a few classic examples here; simplex, tsp.
 
@@
 approaches are just examples in the current lines of research. we
-could also imagine doing [OTHER CREATIVE THING] In classic model,
+could also imagine doing [OTHER CREATIVE THING]In classic model,
 algorithm performance is measured by runtime. In learning we have
 another metric: accuracy. can use smoothing to get results on both
-fronts; see the two examples below.
+fronts;see thetwo examples below.
 
@@
 \subsection{$k$-means is Exponential}
-In \cite{KMworstcase}, Vattani shows that an adversary can choose
-initial cluster centers that will force $k$-means to iterate
-$2^{\Omega(n)}$ times for $d \ge 2$. This construction is based on
-[SOMETHING].
+In \cite{KMworstcase}, Vattani shows that an adversary can choose initial cluster centers that will force $k$-means to iterate $2^{\Omega(n)}$ times for $d \ge 2$.
+This construction is based on [SOMETHING].
 
 \subsection{Smoothed Analysis}
-We give a summary of the argument in \cite{SAkmeans}, by Arthur,
-Manthey, and Roeglin, that $k$-means has polynomial smoothed
-complexity. The main result of this section is the following theorem.
+We give a summary of the argument in \cite{SAkmeans}, by Arthur, Manthey, and Roeglin, that $k$-means has polynomial smoothed complexity. The main result of this section is the following theorem.
 \begin{theorem}
-Fix a set $S^{\prime} \subset [0,1]^d$ of $n$ data points, and
-independently perturb each point in $S^{\prime}$ by a Gaussian
-distribution with mean 0 and variance $\sigma^2$. Label the new set
-$S$. Then the expected running time of $k$-means on $S$ is bounded by
-a polynomial in $n$ and $1/\sigma$.
+Fix a set $S^{\prime} \subset [0,1]^d$ of $n$ data points, and independently perturb each point in $S^{\prime}$ by a Gaussian distribution with mean 0 and variance $\sigma^2$.
+Label the new set $S$.
+Then the expected running time of $k$-means on $S$ is bounded by a polynomial in $n$ and $1/\sigma$.
 \end{theorem}
@@
 \Psi = \sum_{x \in S}||x - c(x)||^2,
-\end{equation} 
-where $c(x)$ is the cluster center to which the point $x$ is assigned
-in the current iteration. The proof relies on two lemmas: first, that
-$\Psi$ is bounded by a polynomial after one iteration (define
-iteration) of the algorithm, and second that $\Psi$ decreases by a
-polynomial amount in each subsequent iteration (actually, sequences of
-a few iterations).
+\end{equation} where $c(x)$ is the cluster center to which the point $x$ is assigned in the current iteration.
+The proof relies on two lemmas: first, that $\Psi$ is bounded by a polynomial after one iteration (define iteration) of the algorithm, and second that $\Psi$ decreases by a polynomial amount in each subsequent iteration (actually, sequences of a few iterations).
 
@@
 \end{proof}
+
+\section{Tensor Decomposition}
+\subsection{Algorithm}
+\subsection{Worst-case}
+\subsection{Smoothed Analysis}
+
+\section{Conclusion}
+
+\begin{thebibliography}{1}
+  \bibitem{SA}
+    Daniel A. Spielman and Shang{-}Hua Teng,
+    ``Smoothed Analysis of Algorithms: Why the Simplex Algorithm Usually takes Polynomial Time.''
+    \emph{Journal of the ACM, Vol 51 (3)},
+    pp. 385 - 463,
+    2004.
+  \bibitem{KMworstcase}
+    Andrea Vattan 
+  \bibitem{SAkmeans}
+    David Arthur, Bodo Manthey, and Heiko Roeglin,
+    ``Smoothed Analysis of the $k$-means Method.''
+    2010.
+\end{thebibliography}
+\end{document}

### 2015-05-14 paper.tex
initial commit

--- 2015-05-14 old.tex
@@
 \noindent Most learning problems are hard in the worst-case, so much
-of the current research focuses on finding special cases with provably
-good time complexities, or polynomial-time approximation algorithms. But many algorithms (Simplex being the
-classical example) seem to consistently run fast
-in practice. Smoothed
-analysis combines worst-case and average-case analysis to better
-approximate real-world algorithm performance and close the gap between
-theoretical and practical results. Here we survey the application of
-smoothed analysis to $k$-means and tensor decomposition, and show that
-both problems admit algorithms with polynomial smoothed complexity, in line with empirical oberservations.
+of the current research focuses on finding good heuristics,
+polynomial-time approximation algorithms, or special cases with
+proveable accuracy and runtime guarantees. But many algorithms that
+are inefficient in the worst case (Simplex being the classical
+example) consistently seem to run fast in practice. Smoothed analysis
+gives a framework for better understanding performance on real-world
+data, specifically for problems where some component is not
+adversarial. This is a natural assumption in learning, since data in
+the learning setting are usually prone to measurement or modeling
+noise. We survey the (very much ongoing) application of smoothed
+analysis to learning problems by way of two examples: $k$-means and
+tensor decomposition.
 \end{abstract}
@@
 \section{Introduction}
--smoothed analysis
-
---what is it? option when theory is far from practice, so good in the
-learning setting. randomly perturb (``smooth'') part of a problem and
-examine what happens on that smoothed instance.give a few classic examples here; simplex, tsp.
-
---what assumptions does it make about the data? when is it useful to
-apply?  when something about problem is not adversarial; i.e. may
-arise randomly in practice.
-
---smoothed analysis in learning. still new; people unsure where to
-apply it: lots of different things to smooth. smooth distribution or
-smooth points? what does it mean to smooth distribution? don't
-know. classically in SA there's one clear thing to smooth but in
-learning we don't really know what's best yet. these 2 papers contain
-different approaches. k-means is runtime-focused. smooths points and
-shows that the runtime on the perturbed set is expected
-polynomial. tensor decomp is accuracy focused. perturbs vectors and
-shows that factorization works robustly with perturbations. these 2
-approaches are just examples in the current lines of research. we
-could also imagine doing [OTHER CREATIVE THING]In classic model,
-algorithm performance is measured by runtime. In learning we have
-another metric: accuracy. can use smoothing to get results on both
-fronts;see thetwo examples below.
-
-% maybe do the main result from the tensor paper instead and mention
-% this result at a higher level. in the conclusion mention how the two
-% papers take different avenues.
-
-\section{$k$-means}
-\subsection{Algorithm}
-The $k$-means algorithm solves the \emph{clustering
-  problem}, set as follows. Given a set $S$ of points in a (generally
-high-dimensional) space $\mathbb{R}^d$, output a $k$-partition of the
-set (a collection of $k$ ``clusters'') such that the elements in each
-cluster are similar under a chosen metric. $k$-means is iterative:
-first it arbitrarily chooses the $k$ cluster centers, then assigns each point in $S$ to the closest center. Next, it
-loops over these clusters and computes the mean of all
-points in the cluster. These means become the new cluster centers and
-the assignment process repeats. The algorithm terminates when the
-local improvements to cluster centers no longer affect the assignment
-of points.
-
-\subsection{$k$-means is Exponential}
-In \cite{KMworstcase}, Vattani shows that an adversary can choose initial cluster centers that will force $k$-means to iterate $2^{\Omega(n)}$ times for $d \ge 2$.
-This construction is based on [SOMETHING].
-
-\subsection{Smoothed Analysis}
-We give a summary of the argument in \cite{SAkmeans}, by Arthur, Manthey, and Roeglin, that $k$-means has polynomial smoothed complexity. The main result of this section is the following theorem.
-\begin{theorem}
-Fix a set $S^{\prime} \subset [0,1]^d$ of $n$ data points, and independently perturb each point in $S^{\prime}$ by a Gaussian distribution with mean 0 and variance $\sigma^2$.
-Label the new set $S$.
-Then the expected running time of $k$-means on $S$ is bounded by a polynomial in $n$ and $1/\sigma$.
-\end{theorem}
-We reason about the potential function \begin{equation}
-\Psi = \sum_{x \in S}||x - c(x)||^2,
-\end{equation} where $c(x)$ is the cluster center to which the point $x$ is assigned in the current iteration.
-The proof relies on two lemmas: first, that $\Psi$ is bounded by a polynomial after one iteration (define iteration) of the algorithm, and second that $\Psi$ decreases by a polynomial amount in each subsequent iteration (actually, sequences of a few iterations).
-
-\begin{lemma}
-The potential after one iteration is bounded by a polynomial.
-\end{lemma}
-\begin{proof}
-\end{proof}
+Spielman and Teng \cite{SA}, \cite{SAtwo} introduced smooth analysis
+to give a more suitable framework for predicting real-life algorithm
+performance. Not every algorithm that runs fast in practice is
+polynomial-time; worst-case analysis falls short of explaining why
+some ``slow'' algorithms are actually quite efficient in practice. The
+original paper \cite{SA} gave a proof that the Simplex algorithm has
+polynomial ``smoothed complexity'': that is, if you assume the data
+are subject to random perturbations (the kind that would arise from
+measurement noise in practice), Simplex runs in expected polynomial
+time. This sparked a host of smoothed analysis proofs involving
+classic combinatorial optimization problems. [EXAMPLES]. The key
+assumption inherent in applying smoothed analysis techniques to an
+algorithm is that some component of the problem is not
+adversarial. The hope is that worst-case inputs are somehow isolated
+in the input space (indeed, the worst-case inputs to many algorithms
+are intricate and fragile), so that any real data will, with high
+probability, not be worst-case. In recent years, there has been an increasing trend of applying smoothed analysis to learning problems \cite{SAtwo}, \cite{SAkmeans}, \cite{PAC}, \cite{TD}.
 
@@
     2004.
+
+  \bibitem{SAtwo}
+    Daniel A. Spielman and Shang{-}Hua Teng,
+    ``Smoothed Analysis: An Attempt to Explain the Behavior of Algorithms in Practice''
+    \emph{Communications of the ACM, Vol. 52 No. 10},
+    pp. 76-84,
+    2009.
+
   \bibitem{KMworstcase}
-    Andrea Vattan 
+    Andrea Vattani,
+    ``k-means Requires Exponentially Many Iterations Even in the Plane.'' 
+    \emph{ Proc. of the 25th ACM Symp. on Computational Geometry (SoCG)}, 
+    pp 324–332, 
+    2009.
+
   \bibitem{SAkmeans}
@@
     2010.
+
+  \bibitem{PAC}
+
+  \bibitem{TD}
+    Aditya Bhaskara, Moses Charikar, Ankur Moitra, and Aravindan Vijayaraghavan,
+    ``Smoothed Analysis of Tensor Decomposition'',
+    CoRR,
+    2013.
+
 \end{thebibliography}

### 2015-05-15 paper.tex
ugh :((((

@@
 polynomial-time approximation algorithms, or special cases with
-proveable accuracy and runtime guarantees. But many algorithms that
-are inefficient in the worst case (Simplex being the classical
-example) consistently seem to run fast in practice. Smoothed analysis
+provable accuracy and runtime guarantees. But many algorithms that are
+inefficient in the worst case consistently seem to run fast in
+practice (Simplex being the classical example). Smoothed analysis
 gives a framework for better understanding performance on real-world
@@
 \section{Introduction}
-Spielman and Teng \cite{SA}, \cite{SAtwo} introduced smooth analysis
+Spielman and Teng \cite{SA}, \cite{SAtwo} introduced smoothed analysis
 to give a more suitable framework for predicting real-life algorithm
@@
 polynomial-time; worst-case analysis falls short of explaining why
-some ``slow'' algorithms are actually quite efficient in practice. The
-original paper \cite{SA} gave a proof that the Simplex algorithm has
-polynomial ``smoothed complexity'': that is, if you assume the data
-are subject to random perturbations (the kind that would arise from
-measurement noise in practice), Simplex runs in expected polynomial
-time. This sparked a host of smoothed analysis proofs involving
-classic combinatorial optimization problems. [EXAMPLES]. The key
-assumption inherent in applying smoothed analysis techniques to an
-algorithm is that some component of the problem is not
-adversarial. The hope is that worst-case inputs are somehow isolated
-in the input space (indeed, the worst-case inputs to many algorithms
-are intricate and fragile), so that any real data will, with high
-probability, not be worst-case. In recent years, there has been an increasing trend of applying smoothed analysis to learning problems \cite{SAtwo}, \cite{SAkmeans}, \cite{PAC}, \cite{TD}.
+some ``slow'' algorithms are empirically quite efficient. As a first
+application of their techniques, the original paper \cite{SA} gave a
+proof that the Simplex algorithm has polynomial ``smoothed
+complexity'': that is, if you assume the data are subject to random
+noice (the kind that would arise in many practical scenarios), Simplex
+runs in expected polynomial time. This sparked a host of papers
+applying smoothed analysis to classic combinatorial optimization
+problems. [EXAMPLES]. The key assumption of the smoothed analysis
+setting is that some component of the problem is not adversarial. The
+hope is that worst-case instances are somehow isolated in the input
+space (indeed, the worst-case inputs to many algorithms are intricate
+and fragile), so that any real data is unlikely to be a worst-case
+instance. In recent years, there has been an increasing trend of
+applying smoothed analysis to learning problems \cite{SAtwo},
+\cite{SAkmeans}, \cite{PAC}, \cite{TD}. Some researchers have followed
+the standard line of smoothed analysis, focusing on algorithm
+\emph{runtimes}. Others have expanded the ideas of smoothed analysis
+to show that algorithms still work under noise. We will explain the
+difference by an example.
+\section{Tensor Decomposition}
+Among other problems, tensor decomposition methods can be used to
+learn topic models, multi-view mixture models, phylogenetic trees, and
+detect communities [CITE--course monograph?]. In practice, tensors we
+want to decompose into factors are random.
 
-\section{Tensor Decomposition}
-\subsection{Algorithm}
-\subsection{Worst-case}
-\subsection{Smoothed Analysis}
-
+\section{$k$-means}
+$k$-means is an algorithm 
 \section{Conclusion}
@@
   \bibitem{PAC}
+    Adam Tauman Kalai, Alex Samorodnitsky, and Shang{-}Hua Teng, 
+    ``Learning and Smoothed Analysis'',
+    \emph{IEEE 54th Annual Symposium on Foundations of Computer Science}, 
+    pp. 395-404,
+    2009.
 

### 2015-05-15 paper.tex
updates

@@
 practice (Simplex being the classical example). Smoothed analysis
-gives a framework for better understanding performance on real-world
-data, specifically for problems where some component is not
+gives a theoretical framework for understanding performance on
+real-world data, specifically for problems where some component is not
 adversarial. This is a natural assumption in learning, since data in
@@
 \end{abstract}
+
+% todo: add formal definition of SA found in \cite{SAtwo} to this section.
 
@@
 proof that the Simplex algorithm has polynomial ``smoothed
-complexity'': that is, if you assume the data are subject to random
-noice (the kind that would arise in many practical scenarios), Simplex
-runs in expected polynomial time. This sparked a host of papers
-applying smoothed analysis to classic combinatorial optimization
-problems. [EXAMPLES]. The key assumption of the smoothed analysis
+complexity'': if you assume the input data are subject to random
+noise, Simplex runs in expected polynomial time. This sparked a host
+of papers applying smoothed analysis to classic combinatorial
+optimization problems. [EXAMPLES]. The key assumption of the smoothed
 setting is that some component of the problem is not adversarial. The
@@
 applying smoothed analysis to learning problems \cite{SAtwo},
-\cite{SAkmeans}, \cite{PAC}, \cite{TD}. Some researchers have followed
-the standard line of smoothed analysis, focusing on algorithm
-\emph{runtimes}. Others have expanded the ideas of smoothed analysis
-to show that algorithms still work under noise. We will explain the
-difference by an example.
+\cite{SAkmeans}, \cite{PAC}, \cite{TD}. (FIX): There are two active fronts:
+some researchers have followed the standard line of smoothed analysis,
+focusing on algorithm \emph{runtimes}. Others have aimed to show that
+algorithms still \emph{work} on perturbed instances of problems; this
+has also been called smoothed analysis, as in \cite{TD}. We give
+examples of both. [explain what 4 and 6 are] In the next section, we
+give a brief formal definition of smoothed analysis, following
+\cite{SAtwo}, to set the environment for the examples that follow. In
+section 3 we discuss at a high level the result of \cite{SAkmeans},
+showing that the $k$-means method has polynomial smoothed runtime. In
+section 4 we examine a tensor decomposition algorithm in a smoothed
+setting, showing not only that it runs in polynomial time on
+smoothed instances, but also that it works provably well, i.e. that it
+recovers good factors. In section 5 we conclude and offer some
+speculation about directions for further research.
+
+\section{Smoothed Analysis: Preliminaries}
+When $A$ is an algorithm for solving problem $P$, we denote by
+$T_A[x]$ the runtime of $A$ on an input instance $x$. We consider the
+input domain $\Omega$ to problem $P$ as the union of disjoint
+subdomains $\{\Omega_1,\ldots, \Omega_n,\ldots\}$, where $\Omega_n$ is
+the family of all inputs of size $n$. Given this setting we can define
+worst-case and smoothed runtimes. The \emph{worst-case measure} is defined as
+
+$$\mbox{WC}_A(n) = \max_{x \in \Omega_n}T_A[x].$$
+The \emph{smoothed complexity} of $A$ with $\sigma$-Gaussian perturbations is given by
+$$\mbox{Smoothed}_A^{\sigma}(n) = \max_{\x\in [-1,1]^n}
+\mathbb{E}_g[T_A(x + g)],$$ where $g$ is a Gaussian random vector of
+variance $\sigma^2$. The original input $x$ is perturbed by $g$ to
+obtain a new input $x + g$, which is given to the algorithm.
+
+We can view $\sigma$ as a parameter that interpolates between
+worst-case and average-case analysis. When $\sigma = 0$ the smoothed
+complexity is the same as the worst-case measure. When $\sigma$ is
+very large, the smoothed complexity approaches the average-case
+complexity, since $g$ dominates the original $x$. Here we are
+interested in the case where $\sigma$ is small compared to $||x||$, so
+that $x + g$ is only a slight perturbation. We now give an example of
+using smoothed analysis to analyze the runtime of a learning
+algorithm, $k$-means.
+
+\section{$k$-means}
+The $k$-means algorithm solves the \emph{clustering problem}, set as
+follows. Given a set $S$ of points in a (generally high-dimensional)
+space $\mathbb{R}^d$, output a $k$-partition of $S$ (a collection of
+$k$ ``clusters'') such that the elements in each cluster are similar
+under a chosen metric. $k$-means begins by arbitrarily choosing the
+$k$ cluster centers (how to choose these centers is the subject of
+much research and debate). The main algorithm consists of two
+iterative steps: first, it assigns each point in $S$ to the closest
+center. Next, it computes the center of mass of each existing cluster
+(the cluster mean). These means become the new cluster centers and the
+assignment process repeats. The algorithm terminates when the local
+improvements to cluster centers no longer affect the assignment of
+points.
+
+In \cite{KMworstcase}, Vattani shows that an adversary can choose
+initial centers that will force $k$-means to iterate $2^{\Omega(n)}$
+times for $d \ge 2$. But $k$-means is usually very fast in practice
+(see e.g. \cite{kmeansfast}). So the setting seems ripe for smoothed
+analysis: there is a large gap between theoretical and observed
+runtimes, and the assumption that data points are subject to noise is
+reasonable. Before we proceed it is important to point out that aside
+from a slow runtime, there is another problem with $k$-means: there
+are no guarantees on the quality of the solution it produces; the
+algorithm is only guaranteed to converge to a local optimum and is
+sensitive to the initial choice of cluster centers
+\cite{kmeanssensitive}. There are mild conditions under which a
+variant of the above algorithm outputs a provably good clustering
+\cite{kmeansprovable}, but these techniques do not use what we
+consider smoothed analysis, instead defining ``separability
+conditions'' on the input instance. We focus on the first problem, a
+problem of runtime, and summarize the argument of Arthur et
+al. \cite{SAkmeans}, the main result of which is the following
+theorem.
+
+\begin{theorem}
+Fix a set $S^{\prime} \subset [0,1]^d$ of $n$ data points, and
+independently perturb each point in $S^{\prime}$ by a Gaussian
+distribution with mean 0 and variance $\sigma^2$. Label the new set
+$S$. Then the expected running time of $k$-means on $S$ is bounded by
+a polynomial in $n$ and $1/\sigma$.
+\end{theorem}
+Again, note that this theorem gives no guarantees on the optimality of
+the output. The argument essentially
+
 \section{Tensor Decomposition}
@@
 learn topic models, multi-view mixture models, phylogenetic trees, and
-detect communities [CITE--course monograph?]. In practice, tensors we
-want to decompose into factors are random.
+detect communities. Tensors are especially useful in learning because
+under mild conditions, the factors of a rank-$k$ 3-tensor decomposition $T =
+\sum_{i=1}^ka_i \otimes b_i \otimes c_i$ are unique
+\cite{tensorunique}.
 
-\section{$k$-means}
-$k$-means is an algorithm 
 \section{Conclusion}
@@
   \bibitem{SA}
-    Daniel A. Spielman and Shang{-}Hua Teng,
+    Daniel A. Spielman and Shang{-}Hua Teng.
     ``Smoothed Analysis of Algorithms: Why the Simplex Algorithm Usually takes Polynomial Time.''
@@
   \bibitem{SAtwo}
-    Daniel A. Spielman and Shang{-}Hua Teng,
+    Daniel A. Spielman and Shang{-}Hua Teng.
     ``Smoothed Analysis: An Attempt to Explain the Behavior of Algorithms in Practice''
@@
   \bibitem{KMworstcase}
-    Andrea Vattani,
+    Andrea Vattani.
     ``k-means Requires Exponentially Many Iterations Even in the Plane.'' 
@@
   \bibitem{SAkmeans}
-    David Arthur, Bodo Manthey, and Heiko Roeglin,
+    David Arthur, Bodo Manthey, and Heiko Roeglin.
     ``Smoothed Analysis of the $k$-means Method.''
@@
   \bibitem{PAC}
-    Adam Tauman Kalai, Alex Samorodnitsky, and Shang{-}Hua Teng, 
+    Adam Tauman Kalai, Alex Samorodnitsky, and Shang{-}Hua Teng.
     ``Learning and Smoothed Analysis'',
@@
   \bibitem{TD}
-    Aditya Bhaskara, Moses Charikar, Ankur Moitra, and Aravindan Vijayaraghavan,
+    Aditya Bhaskara, Moses Charikar, Ankur Moitra, and Aravindan Vijayaraghavan.
     ``Smoothed Analysis of Tensor Decomposition'',
-    CoRR,
+    \emph{CoRR},
     2013.
+
+  \bibitem{tensorunique}
+    J. Kruskal. 
+    ``Three-way Arrays: Rank and Uniqueness of Trilinear Decompositions'',
+    \emph{Linear Algebra and Applications}, 
+    18:95–138, 
+    1977.
+  \bibitem{kmeansprovable}
+    Rafail Ostrovsky and Yuval Rabani and et al.
+    ``The Effectiveness of Lloyd-Type Methods for the k-Means Problem'',
+
+  \bibitem{kmeanssensitive}
+    G. Milligan. 
+    ``An examination of the effect of six types of error perturbation on fifteen clustering
+    algorithms,'' 
+    \emph{Psychometrika}, 
+    45:325–342, 1980.
 

### 2015-05-15 paper.tex
tensorz

@@
 The \emph{smoothed complexity} of $A$ with $\sigma$-Gaussian perturbations is given by
-$$\mbox{Smoothed}_A^{\sigma}(n) = \max_{\x\in [-1,1]^n}
+$$\mbox{Smoothed}_A^{\sigma}(n) = \max_{x\in [-1,1]^n}
 \mathbb{E}_g[T_A(x + g)],$$ where $g$ is a Gaussian random vector of
@@
 \section{Tensor Decomposition}
-Among other problems, tensor decomposition methods can be used to
-learn topic models, multi-view mixture models, phylogenetic trees, and
-detect communities. Tensors are especially useful in learning because
-under mild conditions, the factors of a rank-$k$ 3-tensor decomposition $T =
-\sum_{i=1}^ka_i \otimes b_i \otimes c_i$ are unique
-\cite{tensorunique}.
+(FIX): Among other problems, tensor decomposition methods can be used
+to learn topic models, multi-view mixture models, phylogenetic trees,
+and detect communities. Tensors are especially useful in learning
+because under mild conditions, the factors of a rank-$k$ 3-tensor
+decomposition $T = \sum_{i=1}^ka_i \otimes b_i \otimes c_i$ are unique
+\cite{tensorunique}. These conditions extend naturally to higher-order
+tensors. Assume we are given a $l$-order tensor $T$ of dimension $n$
+(so $T\in \mathbb{R}^{n^{\times l}}$) and of rank $R$ and we want to
+decompose it into a sum of rank-one tensors, i.e.:
+$$T = \sum_{i=1}^R u_i^{(1)}\otimes u_i^{(2)}\otimes\dots\otimes
+u_i^{(l)}$$ We first describe conditions under which such
+decomposition is unique and computable in polynomial time.
+
+The algorithm above works well in the \emph{exact} case, when we know
+$T$. But in the learning setting, we don't know $T$ exactly--we have
+an approximation $\~ T$ of $T$, since we can only take polynomially
+many samples [VAGUE?]. Hence, we need to make sure the algorithm is
+robust in the presence of noise. In particular, we prove that if each
+entry of $E=T-\~ T$ is bounded by $\epsilon \cdot
+\mbox{poly}_l(1/k,1/n,1/\delta)$, then we can compute a decomposition
+in which each rank-$1$ tensor in the decomposition has an additive
+error of at most $\epsilon$.
 

### 2015-05-16 paper.tex
columns?!
inb4 hangry

--- 2015-05-15 paper.tex
@@
 \titlelabel{\thetitle.\quad}
-%\renewcommand\thesection{\Roman{section}} % Roman numerals for the sections
-%\renewcommand\thesubsection{\Roman{subsection}} % Roman numerals for subsections
+\renewcommand\thesection{\Roman{section}} % Roman numerals for the sections
+\renewcommand\thesubsection{\Roman{subsection}} % Roman numerals for subsections
 %\titleformat{\section}[block]{\large\scshape\centering}{\thesection.}{1em}{} % Change the look of the section titles
@@
 \newtheorem{theorem}{Theorem}
+\newtheorem{definition}{Definition}
 
@@
 of papers applying smoothed analysis to classic combinatorial
-optimization problems. [EXAMPLES]. The key assumption of the smoothed
+optimization problems \cite{SAtwo}. The key assumption of the smoothed
 setting is that some component of the problem is not adversarial. The
@@
 applying smoothed analysis to learning problems \cite{SAtwo},
-\cite{SAkmeans}, \cite{PAC}, \cite{TD}. (FIX): There are two active fronts:
-some researchers have followed the standard line of smoothed analysis,
-focusing on algorithm \emph{runtimes}. Others have aimed to show that
-algorithms still \emph{work} on perturbed instances of problems; this
-has also been called smoothed analysis, as in \cite{TD}. We give
-examples of both. [explain what 4 and 6 are] In the next section, we
-give a brief formal definition of smoothed analysis, following
+\cite{SAkmeans}, \cite{PAC}, \cite{TD}.
+The noisy assumptions required in the smoothed model are natural in the learning setting, where data are gathered by surveys, measurements, samples, and other noise-prone or imperfect methods.
+In the next section, we give a brief formal definition of smoothed analysis, following
 \cite{SAtwo}, to set the environment for the examples that follow. In
-section 3 we discuss at a high level the result of \cite{SAkmeans},
-showing that the $k$-means method has polynomial smoothed runtime. In
-section 4 we examine a tensor decomposition algorithm in a smoothed
+Section III we discuss at a high level the result of \cite{SAkmeans},
+showing that the $k$-means clustering method has polynomial smoothed runtime. In
+Section IV we examine a tensor decomposition algorithm in a smoothed
 setting, showing not only that it runs in polynomial time on
 smoothed instances, but also that it works provably well, i.e. that it
-recovers good factors. In section 5 we conclude and offer some
+recovers good factors. In Section V we conclude and offer some
 speculation about directions for further research.
 
-\section{Smoothed Analysis: Preliminaries}
+\section{Preliminaries}
 When $A$ is an algorithm for solving problem $P$, we denote by
@@
 times for $d \ge 2$. But $k$-means is usually very fast in practice
-(see e.g. \cite{kmeansfast}). So the setting seems ripe for smoothed
+(see, for instance, \cite{kmeansfast}). So the setting seems ripe for smoothed
 analysis: there is a large gap between theoretical and observed
@@
 reasonable. Before we proceed it is important to point out that aside
-from a slow runtime, there is another problem with $k$-means: there
+from a slow worst-case runtime, there is another problem with $k$-means: there
 are no guarantees on the quality of the solution it produces; the
@@
 \begin{theorem}
+\label{sakmeans}
 Fix a set $S^{\prime} \subset [0,1]^d$ of $n$ data points, and
@@
 \end{theorem}
-Again, note that this theorem gives no guarantees on the optimality of
-the output. The argument essentially
+The proof of Theorem \ref{sakmeans} makes heavy use of the potential
+function $$\Psi = \sum_{x \in S}||x - c(x)||^2,$$ where $c(x)$ is the
+center of the cluster to which the point $x$ is assigned, and consists
+of two steps. It first shows that the potential is bounded by a
+polynomial after the first iteration of the algorithm. The second step
+shows that the potential decreases by a polynomial in each iteration
+(or rather, in a sequence of iterations). The key insight of the paper
+is its introduction of \emph{transition blueprints}, directed
+multigraphs that model the flow of points between clusters in an
+iteration of the algorithm. Vertices are clusters and edges are drawn when points in $S$ go from one cluster to another in an iteration. The potential can decrease in two ways:
+cluster centers can change, or points can be reassigned from one
+cluster to another. No iteration can decrease the potential. For the
+potential difference to be small in a given iteration, it must be the
+case that no cluster center moved very much, and that no points were
+reassigned to a cluster very far from their current cluster. The paper
+examines the probability that the potential decrease is small for a
+given transition blueprint; this leads naturally to an upper-bound on
+the probability that the minimum potential decrease across all iterations is small.
+Because of the small perturbations of the input set, transition blueprints are general enough so that each one represents multiple possible iterations of the algorithm, but specific enough that they \emph{almost} uniquely determine what happens during an iteration.
+The paper does a case-by-case analysis of different types of blueprints to show that with high probability, there are no iterations where cluster centers don't move far and points don't get swapped between far-away clusters.
+This implies the result.
+
+$k$-means has expected polynomial smoothed complexity, matching observations
+in practice. The full analysis is quite intricate, so the resulting
+polynomial has degree much higher than empirical results would
+suggest, but this is in line with other results in smoothed analysis
+of runtimes \cite{SAtwo}. But we still have no provable guarantees for
+optimality. In the next section, we examine in detail the more interesting example of analyzing an algorithm
+\emph{with provable guarantees} in a smoothed setting.
 
 \section{Tensor Decomposition}
-Among other problems, tensor decomposition methods can be used to
-learn topic models, multi-view mixture models, phylogenetic trees, and
-detect communities. Tensors are especially useful in learning because
-under mild conditions, the factors of a rank-$k$ 3-tensor decomposition $T =
-\sum_{i=1}^ka_i \otimes b_i \otimes c_i$ are unique
-\cite{tensorunique}.
+Tensor decomposition methods can be used for community detection and for learning  topic models (including LDA), multi-view mixture models, and phylogenetic trees.
+Tensors are especially useful because
+under mild conditions, the factors of a rank-$R$ 3-tensor decomposition $T =
+\sum_{i=1}^Ra_i \otimes b_i \otimes c_i$ are unique
+\cite{tensorunique}, and there exists a polynomial-time algorithm for recovering them, initially due to Jennrich.
+Denote by $A$ the $n\times R$ matrix whose columns are the $a_i$, and likewise for $B$.
+Then the conditions for Jennrich's algorithm are that $A$ and $B$ have rank $R$, and that no two vectors $c_i, c_j$ are linearly dependent.
+Jennrich's algorithm extends naturally to higher-order tensors.
+Assume we are given a order-$l$ tensor $T$ of dimension $n$ (so $T\in \mathbb{R}^{n^{\times l}}$) and rank $R$.
+We want to decompose it into a sum of rank-one tensors, i.e.:
+$$T = \sum_{i=1}^R u_i^{(1)}\otimes u_i^{(2)}\otimes\dots\otimes u_i^{(l)}.$$
+Jennrich's algorithm works for order-$3$ tensors, so to work with the general case a good approach is to flatten a tensor $T$ into an order $3$ tensor so as to obtain a decomposition, through Jennrich's, of the form
+$$T = \sum_{i=1}^R U_i\otimes V_i\otimes W_i$$
+where
+$$U_i=u_i^{(1)}\odot\dots\odot u_i^{(\lfloor\frac{l-1}{2}\rfloor)},\  V_i=v_i^{(\lfloor\frac{l-1}{2}\rfloor+1)}\odot\dots\odot v_i^{(2\lfloor\frac{l-1}{2}\rfloor)} \text{, and } W_i = u_i^{(l)}\text{ or }u_i^{(l-1)}\odot u_i^{(l)}.$$
+Here $\odot$ is the Khatri-Rao product: given $M \in \mathbb{R}^{m\times r},\ N \in \mathbb{R}^{n\times r}$, $M\odot N$ is the matrix of dimension $mn\times r$ whose $i$-th column is $M_i\otimes N_i$.
+[MORE--it also only works for 3-tensors] Jennrich's algorithm works only if $R \le n$, that is, only if the rank of the tensor is at most the dimension of its factors.
+In the overcomplete case where $R > n$, we can "flatten" the factors using the Khatrio-Rao product; if the flattened factors still satisfy the conditions of Jennrich's algorithm, then we can solve this higher-order decomposition problem.
+Also note that overcomplete decomposition is NP-hard in general \cite{tensorhard1} \cite{tensorhard2}.
+[MENTION REM 1.8 before next sentence / paragraph] We show that in a smoothed setting where the tensor $T$ is perturbed, the flattened factors will indeed satisfy the conditions for Jennrich's algorithm, so we can get around the worst-case inefficiency.
 
+TODO: put remark 1.8, condition 2.2 and a noiseless version of corollary 2.4 here. they explain very clearly why we want to show that the kruskal rank multiplies.
+
+The overcomplete algorithm above [THE ONE MENTIONED IN REM 1.8] works well in the \emph{exact} case, when we know $T$.
+But in the learning setting, we don't know $T$ exactly--we have an approximation $\~ T$ of $T$, since we can only take polynomially many samples [VAGUE?].
+Hence, we need to make sure the algorithm is robust in the presence of noise.
+In particular, we prove that if each entry of $E=T-\~ T$ is bounded by $\epsilon \cdot \mbox{poly}_l(1/k,1/n,1/\delta)$, then we can compute a decomposition in which each rank-$1$ tensor in the decomposition has an additive error of at most $\epsilon$.
+
+%technical lemma
+\begin{lemma}
+Let $\mathcal{V}\subset \mathbb{R}^{nm}$ be a subspace with $\dim(\mathcal{V})=\delta nm$.
+Furthermore, suppose $r,\theta,\delta'$ satisfy $\delta'\leq \delta/2$, $r\delta' m<\delta n/2$ and $\theta= \frac{1}{nm^{3/2}}$.
+Then there exist $r$ matrices $M_1, M_2,\dots, M_r$ of dimension $n\times m$ such that
+\begin{enumerate}
+\item $vec(M_i)\in \mathcal{V}$ $\forall i\in [r]$
+\item $M_1,\dots, M_r$ form an ordered $(\theta,\delta')$ orthogonal system
+
+\end{enumerate}
+where $\vec{M_i}$ is the vector resulting from flattening $M_i$. In particular, if $m\leq \sqrt{n}$, the matrices form an ordered $(\theta, \delta/2)$ orthogonal system.
+\end{lemma}
+\begin{proof}
+We treat vectors $M\in \mathcal{V}$ as $n\times m$ matrices for simplicity. The proof of Lemma 1 involves various intermediate results.
+
+%proof of technical lemma
+
+\begin{definition} (Robust Dimension of Projections) Let
+$$\dim_i^\tau (\mathcal{V}) = \max_{d}\text{s.t. }\exists\text{ orthonormal }v_1,\dots,v_d\in \mathbb{R}^n \text{ and }M_1,\dots,M_d\in\mathcal{V} $$
+$$\text{ with } ||M_t||\leq \tau \text{ and } v_t=M_t(i)\text{ }\forall t\in[d]$$
+\end{definition}
+\noindent We try to find many column projections having robust dimension of about $\delta n$. We have the
+
+\begin{lemma}
+For a subspace $\mathcal{V}\subset \mathbb{R}^{p_1\cdot p_2}$ and for any $\tau\geq \sqrt{p_2}$ we have:
+$$\sum_{i\in[p_2]}\dim_i^\tau (\mathcal{V})\geq \dim(\mathcal{V})$$
+\end{lemma}
+
+\begin{proof}
+Writing $d=\dim(\mathcal{V})$ and taking $B$ to be a $(p_1p_2)\times d$ matrix whose columns are an orthonormal basis of $\mathcal{V}$ we have $\sigma_d(B)=1$.
+We form $p_2$ matrices of dimension $p_1$ by indexing the rows of $B$ by $[p_1]\times[p_2]$ and taking $B_i$ to be the rows corresponding to indices of the form $(j,i)$ with $j\in[p_1]$.
+We define
+$d_i=\max_t \text{s.t. } \sigma_t(B_i)\geq \frac{1}{\sqrt{p_2}}$
+We show $$\sum_{i\in[p_2]}\dim_i^\tau (\mathcal{V})\geq \sum_{i\in[p_2]} d_i\geq \dim(\mathcal{V})$$.
+
+Suppose the RHS of the inequality was false.
+Then define $S_i\subset \mathbb{R}^d$ to be the $d-d_i$-dimensional subspace spanned by the $d-d_i$ rightmost singular vectors of $B_i$.
+Since $d-\sum_i d_i>0$, $\cap_i S_i^\perp\neq \emptyset$, so we can take $\alpha\in \cap_i S_i^\perp$ to be a unit vector.
+Then for every $i$ we have $||B_i\alpha||\leq \frac{1}{\sqrt{p_2}}$, but this implies $\sigma_d(B)\leq \sum_i||B_i\alpha||^2\leq p_2\cdot\frac{1}{p_2}=1$, a contradiction.
+
+For the LHS, we consider the $d_i$ top left-singular vectors of $B_i$.
+From a general result for matrices (see lemma B.1 in \cite{TD}), they can be represented as $\beta_j B_i$ where $1\leq j\leq d_i$ and $\beta_j\in \mathbb{R}^{d}$ satisfies $||\beta_j||_2\leq \sqrt{p_2}$.
+Then $\beta_1B,\dots,\beta_{d_i}B$ are vectors in $\mathbb{R}^{p_1p_2}$ that show $\dim_i^\tau(\mathcal{V})\geq d_i$.
+
+\end{proof}
+
+We can now construct the matrices $M_i$
+
+%end of proof
+\end{proof}
 \section{Conclusion}
+Smoothed analysis is well-suited to learning problems because learning
+data are inherently noisy. What's not so clear is the best way to
+apply these techniques: what exactly should be ``smoothed'' with
+perturbations? What should we look at once we've perturbed the inputs?
+In this survey we've given two examples of smoothed analysis in
+learning: one in a setting with provable guarantees and the other
+without. These two examples represent, to us, two different vectors
+for further research. Many learning heuristics have no good runtime
+bounds but run efficiently in practice. In these cases, smoothed
+analysis may be able to bridge the gap, as it has for $k$-means. Separately, there are several machine learning
+algorithms that work provably well under certain mild input
+conditions--finding such algorithms has been a major subject of
+machine learning research in recent years. Less well-known is whether these
+input conditions are preserved by the application of noise, and if
+they are, whether the output of the algorithm is stable. In the case
+of tensor decomposition, \cite{TD} indeed shows that the conditions
+for uniqueness are preserved by Gaussian perturbation, and that the
+factors of the perturbed tensor are close to the factors of the
+original tensor, so the output is consistent. Whether these results translate to other algorithms seems to be a wide-open area for further investigation.
 
@@
     \emph{Communications of the ACM, Vol. 52 No. 10},
-    pp. 76-84,
+    pp. 76 - 84,
     2009.
@@
     \emph{ Proc. of the 25th ACM Symp. on Computational Geometry (SoCG)}, 
-    pp 324–332, 
+    pp 324-332, 
     2009.
@@
     \emph{IEEE 54th Annual Symposium on Foundations of Computer Science}, 
-    pp. 395-404,
+    pp. 395 - 404,
     2009.
@@
     \emph{Linear Algebra and Applications}, 
-    18:95–138, 
+    18:95 - 138, 
     1977.
@@
     \emph{Psychometrika}, 
-    45:325–342, 1980.
+    45:325 - 342, 1980.
 
+  \bibitem{kmeansfast}
+    Sariel Har{-}Peled and Bardia Sadri.
+    ``How fast is the k-means method?'',
+    \emph{Algorithmica},
+    41(3):185 - 202,
+    2005.
+
+  \bibitem{tensorhard1}
+    Johan H{\aa}stad.
+    ``Tensor Rank is NP-complete'',
+    \emph{Journal of Algorithms}
+    pp. 644 - 654,
+    1990.
+  \bibitem{tensorhard2}
+    Christopher J. Hillar and Lek{-}Heng Lim.
+    ``Most tensor problems are {NP} hard''
+    \emph{CoRR},
+    2009.
 \end{thebibliography}
